---
date: |
  `r format(Sys.time(), '%d %b %Y')`
title: | 
  Chapter 2 draft: something something working lands under public and private control have equal grassland bird community whatevers
author: 
  - Megan Massa, University of Maryland Center for Environmental Science, 301 Braddock Road, Frostburg, MD 21532, USA
  # - Thomas L. Serfass, Frostburg State University, 101 Braddock Road, Frostburg, MD 21532, USA
  # - Amy E. M. Johnson, Smithsonian Conservation Biology Institute, 1500 Remount Road, Front Royal, VA 22630, USA
  # - Elizabeth R. Matthews, National Park Service, 1100 Ohio Drive SW, Washington, DC 20242, USA
  # - W. Gregory Shriver, University of Delaware, 210 S College Avenue, Newark, DE 19716, USA
  # - Emily B. Cohen, University of Maryland Center for Environmental Science, 301 Braddock Road, Frostburg, MD 21532, USA
output: 
  bookdown::word_document2:
    reference_docx: 'template_simple.docx'
    number_sections: false
    linenumbers: yes
  bookdown::html_document2:
    toc: true
    toc_float:
      collapsed: true
    toc_depth: 3
    fig_caption: true
    number_sections: false
bibliography: ./Thesis.bib
csl: ./condor_style.csl
link-citations: yes
---
```{r setup, include = F}

# Don't show code chunks
knitr::opts_chunk$set(message = F, warning = F, echo = F)

# Load required packages
library(knitr)
library(kableExtra)
library(flextable)
library(tidyverse)
library(ggpubr)
library(unmarked)
source('scripts/wrangling/tidy_method_unmarked.R')

# Don't show NAs in kables
options(knitr.kable.NA = '')

# If simple template, Flextable defaults to Calibri 11. Otherwise, TNR 12
if (rmarkdown::metadata$output$`bookdown::word_document2`$reference_docx == 'template_simple.docx')
  set_flextable_defaults(font.family = 'Calibri', font.size = 11) else
    set_flextable_defaults(font.family = 'Times New Roman', font.size = 12) 

# Set plot theme
theme_set(
  theme_bw() +
  theme(
    strip.text = element_text(size = 11),
    axis.text = element_text(size = 9),
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 10),
    panel.grid = element_blank(),
    strip.background = element_blank()))
```

*Correspondence*: Megan Massa, University of Maryland Center for Environmental Science, 301 Braddock Road, Frostburg, MD 21532, USA. Email: megan.massa@umces.edu

**KEYWORDS** `r paste(sort(c("grassland birds", "protected areas", "working lands")), collapse = ", ")`

\newpage

# Abstract

<!-- The Abstract is perhaps the most important paragraph in a paper. Avoid writing it as a mere summary of the research. Place key results in the abstract’s first two sentences – don't bury the lede. Do not include lists of common methods, discursive explanations of what you set out to accomplish, or citations. Use the abstract and title to highlight the paper’s keywords. Maximum word count is 350.  -->

# Introduction

<!-- This section provides your research aims and significance and places it within the framework of the existing literature. The last Introduction paragraph typically provides the study’s hypotheses and testable predictions. Use citations judiciously; most points can be supported by three or fewer citations. Long lists of citations are seldom required and detract from the readability of the manuscript. -->

Biodiversity is in decline. 
Avian biodiversity in particular.
Particularly grasslands.
Protected areas are an important way to maintain biodiversity.
But working lands not under strict protection are also important.

Summarize research on community index (forest focused) and role of protected areas (not focused on grassland birds in this region).
Role of easements.

We surveyed bird communities in Virginia, Maryland, and West Virginia to evaluate the relative contributions of protected areas and private land for grassland bird conservation.
Specifically, we examined bird populations in four national battlefield parks and on XXX properties enrolled in Virginia Working Landscapes
We hypothesized that protected areas would host higher grassland bird richness and diversity, and that 

# Methods

<!-- Provide enough information to enable readers to replicate and critically evaluate the research. Describe statistical tests and procedures. Cite statistical software and analysis programs. Explain how summary statistics are reported in Results, e.g., means ± 1 SE.  -->

The NPS I&M has collected forest bird occurrence and abundance data in the National Capital Region since 2007 [@nationalparkservice2005] and in 2014 initiated long-term population monitoring for grassland birds.
Virginia Working Landscapes

We classified species as obligate and facultative grassland-breeders following @vickery1999 Table \@ref(tab:species)). 

Calculated richness and Shannon diversity of grassland species using the R package "vegan" [@oksanen2022]

Modeled

Although abundance can be modeled as a hierarchical process

We performed all analyses with program R version 4.1.2 [@rcoreteam2022]. 

```{r species}
read_rds("data/processed/sp_grs_ch2.rds") %>% 
  select(
    `Common name` = common_name,
    `Scientific name` = sci_name,
    Designation = status) %>% 
  flextable() %>% 
  set_caption(
    "Grassland birds detected. Designation from Vickery et al. (1999).") %>% 
  theme_booktabs(bold_header = TRUE) %>% 
  autofit() %>% 
  fit_to_width(max_width = 6.5) %>% 
  italic(j = 2) %>% 
  line_spacing(space = 0.5)
```

# Results

<!-- Include only results pertinent to the hypotheses being tested or questions raised in the Introduction section and treated in the Discussion. Use the same number of decimal places for means and estimates of uncertainty (e.g., 38.9 ± 1.2, not 38.9 ± 1.23); usually only one or two decimal places are necessary. Round percentages to whole numbers. The text should not duplicate material presented in tables or figures. Clearly present relevant sample sizes, degrees of freedom, values of statistical tests, and P-values. Test statistics should be rounded to one (t-test, C2, F, etc.) or two (r, r2, etc.) decimal places. See Statistical Details in the Style Sheet for more information. -->


# Discussion

<!-- Begin with a statement that summarizes the study’s main findings. Develop the significance and importance of the results and set them into a framework of previous research. Additional statistical tests and results are inappropriate here except in unusual cases. If you have a Conclusion section, list it as a subheading within the Discussion. -->


# Acknowledgments

Thank sponsors, everyone who helped with research, and reviewers

**Funding statement**

**Ethics statement**

**Conflict of interest statement**

**Author contributions**

**Data availability**

# Tables

# Figures

# References
