---
title: "Field season summary"
date: "Generated `r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
options(knitr.table.format = "html")
options(knitr.kable.NA = '')
library(sf)
library(tmap)
library(lubridate)
library(tidyverse)
library(kableExtra)
library(DT)

report_year <- 2018

theme_set(theme_bw())
```

```{r header_image, fig.align='center', out.width= '50%'}
# knitr::include_graphics(
#   path = 'https://upload.wikimedia.org/wikipedia/commons/thumb/3/32/Bull_Run_at_Manassas_National_Battlefield_Park_09.JPG/1024px-Bull_Run_at_Manassas_National_Battlefield_Park_09.JPG')
```

**NPS Inventory and Monitoring, National Capital Region Network** 

Document created by Megan Massa 

University of Maryland Center for Environmental Science, Appalachian Laboratory

<hr>

# {.tabset}

## `r report_year` Summary

This document summarizes survey efforts and results for grassland bird monitoring during the `r report_year` field season.

```{r data_import, include = FALSE}
# Import tidied data (see import_bird_data.R)
read_rds('data/processed/birds.rds') %>% 
  list2env(.GlobalEnv)

visits <- filter(visits, year == report_year)

counts <- filter(counts, visit_id %in% visits$visit_id)

# Create summary: number of visits
n_visits <- 
  visits %>% 
  group_by(grts) %>% 
  summarize(visits = length(unique(visit_id))) %>% 
  group_by(visits) %>% 
  summarize(n_points = n())

# Create variables for summary paragraph
total_pts <- colSums(n_visits)[2]
three_visits <- 
  if(any(n_visits$visits == 3)) {unlist(filter(n_visits, visits == 3)[2])} else {0}
two_visits <- 
  if(any(n_visits$visits == 2)) {unlist(filter(n_visits, visits == 2)[2])} else {0}
one_visit <- 
  if(any(n_visits$visits == 1)) {unlist(filter(n_visits, visits == 1)[2])} else {0}
date_min <-
  min(visits$date) %>% format('%d %b')
date_max <-
  max(visits$date) %>% format('%d %b')
time_min <-
  round(abs(min(visits$start_sun)))
time_max <-
  round(abs(max(visits$start_sun)))
```

### Monitoring effort

The bird monitoring team conducted 
`r count(filter(visits, year == report_year))` 
surveys at 
`r total_pts` 
total points, of which 
`r three_visits`
`r if_else(three_visits == 1, 'point was', 'points were')`
visited three times,
`r two_visits` 
`r if_else(two_visits == 1, 'point was', 'points were')` 
visited only twice and 
`r one_visit` 
`r if_else(one_visit == 1, 'point was', 'points were')` 
visited only once. 
Surveys ran from `r date_min` to `r date_max` and were conducted from `r time_min` minutes before local sunrise to `r time_max` minutes (`r round(time_max / 60, 2)` 
hours) after local sunrise. 

```{r observer_table}
visits %>% 
  group_by(observer) %>% 
  summarize(surveys = n()) %>% 
  kable(col.names = c('Observer', 'Surveys')) %>% 
  kable_styling(full_width = FALSE)

```

### Survey conditions

The following graphs summarize environmental conditions during surveys. Dotted red lines indicate threshold recommended values according to survey protocol; surveys past these lines were not conducted under advised conditions.

```{r survey_conditions1, fig.show = "hold", fig.asp = 0.25}
# Disturbance
ggplot(visits, aes(x = disturbance)) + geom_bar() + 
  scale_x_continuous(breaks = seq(from = 0, to = max(visits$disturbance), by = 1))

# Wind
ggplot(visits, aes(x = wind)) + geom_bar() +
  scale_x_continuous(breaks = seq(from = 0, to = max(visits$wind), by = 1)) +
  geom_vline(xintercept = 3.5, color = "red", lty = 2)

# Sky
ggplot(visits, aes(x = sky)) + geom_bar()

# Start time
ggplot(visits, aes(x = start_sun)) + 
  geom_histogram(binwidth = 2.5, closed = "left") + 
  geom_vline(xintercept = -20, color = "red", lty = 2) + 
  geom_vline(xintercept = 300-10, color = "red", lty = 2) +
  labs(x = "start time: minutes since sunrise")

# Date
ggplot(visits, aes(x = date)) + geom_histogram(binwidth = 1) 

# Temperature
ggplot(visits, aes(x = temperature)) + geom_histogram(bins = 30)

# Humidity
ggplot(visits, aes(x = humidity)) + geom_histogram(bins = 30)
```

```{r survey_conditions2, fig.asp = 0.7}
# Temp vs humid
ggplot(visits, aes(x = humidity, y = temperature)) + geom_point() 
```


### Bird detections

```{r species_summary}
taxonomy <-
  read_rds('data/processed/taxonomy.rds')

species_list <-
  distinct(counts, species) %>% 
    left_join(taxonomy) %>% 
    left_join(
      readxl::read_xlsx('data/raw/ACAD Global 2021.02.05.xlsx', 1) %>% 
        select(common_name = `Common Name`, 
               habitat1 = `Primary Breeding Habitat`, 
               habitat2 = `Secondary Breeding Habitat`,
               iucn = `IUCN Red List 2018`,
               pif = `CI`),
      by = "common_name") %>% 
    arrange(tax_order)

full_data <-
  counts %>% 
  left_join(visits) %>% 
  left_join(points) %>% 
  left_join(species_list)

species_by_park <-
  full_data %>% 
  group_by(park) %>% 
    select(any_of(names(species_list))) %>% 
  group_by(park, species) %>% 
  summarize(detections = n()) %>% 
  left_join(species_list) %>% 
  arrange(park, tax_order) %>% 
  ungroup()

park_summary <-
  full_data %>% 
  group_by(park) %>% 
  summarize(
    surveys = n_distinct(visit_id),
    points = n_distinct(grts),
    species = n_distinct(species[tax_level=='species']))

grass_spp <- 
  species_list %>% 
  filter(str_detect(habitat1, "Grassland") | str_detect(habitat2, "Grassland")) %>% 
  pull(species)

# All obligate grassland species of temperate north america from Vickery et al. 1999
vickery <-
  c("NOHA","SWHA","FEHA","RLHA","STGR","GRPC","LEPC","MOPL","UPSA","LBCU","MAGO","BUOW","LEOW","SEOW","HOLA","SEWR","AMPI","SPPI","BACS","VESP","LABU","SAVS","GRSP","BAIS", "HESP","LCSP","TBLO","CCLO","DICL","BOBO","EAME","WEME")

# All facultative grassland species of temperate north america from Vickery et al. 1999
vickery2 <-
  c('AMBI','CAEG','TUVU','CANG','GADW','AMWI','MALL','BWTE','NSHO','NOPI','GWTE','AMKE','MERL','PEFA','PRFA','RNEP','WIPT','NOBO','YERA','SACR','WHCR','KILL','LEYE','WILL','LESA','WISN','WIPH','FRGU','MODO','BANO','CONI','COPO','SAPH','ATFL','CAKI','WEKI','EAKI','STFL','LOSH','NSHR','CHRA','EABL','WEBL','MOBL','COYE','CANT','RWSP','RCSP','CCSP','LASP','RWBL','BRBL','BRCO','BHCO')

vickery_spp <- filter(species_list, species %in% c(vickery, vickery2))$species

cbsd_spp <- filter(species_list, pif == "CBSD")$species

yellow_spp <- filter(species_list, pif == "Watch List - Yel-D")$species
  
```

In `r report_year` there were `r species_list %>% filter(tax_level == 'species') %>% nrow()` bird species detected during grassland surveys. There were `r n_distinct(c(grass_spp, vickery_spp))` grassland species (defined as species that either obligately or facultatively breed in grasslands, as specified by Vickery (1999) and the Partners in Flight (PIF) Avian Conservation Assessment Database) detected, `r length(cbsd_spp)` species ranked by PIF as Common Birds in Steep Decline, and `r length(yellow_spp)` species placed on the PIF Yellow Watch List D for steep declines and major threats.

```{r park_table}
kable(
  park_summary,
  col.names = c('Park', 'Surveys', 'Points', 'Species')) %>% 
  kable_styling(full_width = FALSE)
```

#### Grassland species by park {.tabset}

The following tables show all species designated as obligate or facultative grassland species, along with all species of concern designated by Partners in Flight. Gold-highlighted species are considered grassland obligates and light yellow highlighted species are facultative grassland species (Vickery 1999).

##### ANTI

Grassland species detected at Antietam National Battlefield in `r report_year`:
```{r ANTI_species}
species_ANTI <-
  species_by_park %>% 
  filter(park == 'ANTI', 
         species %in% c(grass_spp, cbsd_spp, yellow_spp, vickery_spp)) %>% 
  select(species, common_name, sci_name, 
         tax_level, tax_order, detections, pif) %>% 
  arrange(-detections, tax_order)

species_ANTI %>% 
  kable(
    col.names = c('Alpha code', 'Common name', 'Scientific name', 
                  'Taxonomic level', 'Taxonomic #', 'Detections', "Watchlist")) %>% 
  kable_styling(full_width = FALSE) %>% 
  row_spec(
    which(species_ANTI$species %in% vickery),
    background = 'gold') %>% 
  row_spec(
    which(species_ANTI$species %in% vickery2),
    background = '#fdffc7') %>% 
  column_spec(3, italic = TRUE) %>% 
  column_spec(2, bold = TRUE) %>% 
  remove_column(4:5)
```

##### HAFE
Species of interest detected at Harpers Ferry National Historical Park in `r report_year`:
```{r HAFE_species}
species_HAFE <-
  species_by_park %>% 
  filter(park == 'HAFE', 
         species %in% c(grass_spp, cbsd_spp, yellow_spp, vickery_spp)) %>% 
  select(species, common_name, sci_name, 
         tax_level, tax_order, detections, pif) %>% 
  arrange(-detections, tax_order)

species_HAFE %>% 
  kable(
    col.names = c('Alpha code', 'Common name', 'Scientific name', 
                  'Taxonomic level', 'Taxonomic #', 'Detections', "Watchlist")) %>% 
  kable_styling(full_width = FALSE) %>% 
  row_spec(
    which(species_HAFE$species %in% vickery),
    background = 'gold') %>% 
  row_spec(
    which(species_HAFE$species %in% vickery2),
    background = '#fdffc7') %>% 
  column_spec(3, italic = TRUE) %>% 
  column_spec(2, bold = TRUE) %>% 
  remove_column(4:5)
```

##### MANA
Species of interest detected at Manassas National Battlefield Park in `r report_year`:
```{r MANA_species}
species_MANA <-
  species_by_park %>% 
  filter(park == 'MANA', 
         species %in% grass_spp | species %in% cbsd_spp | species %in% yellow_spp) %>% 
  select(species, common_name, sci_name, 
         tax_level, tax_order, detections, pif) %>% 
  arrange(-detections, tax_order)

species_MANA %>% 
  kable(
    col.names = c('Alpha code', 'Common name', 'Scientific name', 
                  'Taxonomic level', 'Taxonomic #', 'Detections', "Watchlist")) %>% 
  kable_styling(full_width = FALSE) %>% 
  row_spec(
    which(species_MANA$species %in% vickery),
    background = 'gold') %>% 
  row_spec(
    which(species_MANA$species %in% vickery2),
    background = '#fdffc7') %>% 
  column_spec(3, italic = TRUE) %>% 
  column_spec(2, bold = TRUE) %>% 
  remove_column(4:5)
```

##### MONO
Species of interest detected at Monocacy National Battlefield in `r report_year`:
```{r MONO_species}
species_MONO <-
  species_by_park %>% 
  filter(park == 'MONO',
         species %in% grass_spp | species %in% cbsd_spp | species %in% yellow_spp) %>% 
  select(species, common_name, sci_name, 
         tax_level, tax_order, detections, pif) %>% 
  arrange(-detections, tax_order)

species_MONO %>% 
  kable(
    col.names = c('Alpha code', 'Common name', 'Scientific name', 
                  'Taxonomic level', 'Taxonomic #', 'Detections', "Watchlist")) %>% 
  kable_styling(full_width = FALSE) %>% 
  row_spec(
    which(species_MONO$species %in% vickery),
    background = 'gold') %>% 
  row_spec(
    which(species_MONO$species %in% vickery2),
    background = '#fdffc7') %>% 
  column_spec(3, italic = TRUE) %>% 
  column_spec(2, bold = TRUE) %>% 
  remove_column(4:5)
```

### Grassland species richness by point {.tabset}

```{r point_summary}
point_sf <-
  full_data %>% 
    group_by(grts) %>% 
    summarize(
      species = length(unique(species)),
      visits = length(unique(visit_id)),
      mean_disturb = mean(disturbance, na.rm = TRUE)) %>% 
    left_join(points) %>% 
    st_as_sf(coords = c('long', 'lat'), crs = 4326)
```

The total species detected at each point in `r report_year` is mapped below. Points are outlined with the mean disturbance recorded at the point that year.

```{r point_map_setup, include = FALSE}
park_lines <-
  st_read('data/processed/focal_parks.shp') %>% 
    st_transform(crs = 4326) %>% 
    st_make_valid()

make_map <-
  function(park){
  
  park_boundary <-
    filter(park_lines, UNIT_CODE == park) %>% 
    st_transform(3857)
  
  park_point_sf_large <-
    filter(point_sf, park == park) %>% 
    st_transform(3857) %>% 
    st_buffer(dist = 125, nQuadSegs = 40)
  
  park_point_sf_small <-
    filter(point_sf, park == park) %>% 
    st_transform(3857) %>% 
    st_buffer(dist = 100, nQuadSegs = 50)
  
  tmap_options(check.and.fix = TRUE)
  
  richness_map <- 
    tm_basemap('Esri.WorldImagery') +
        
    # SAMPLING UNITS
    tm_shape(park_boundary) +
    tm_borders(
      col = 'yellow',
      lwd = 2.5) +
    
      tm_shape(park_point_sf_large) +
      tm_fill(
        col = 'mean_disturb',
        style = 'cont',
        palette = 'Greys',
        id = 'point_name',
        popup.vars = 
          c('Point name' = 'point_name',
            'GRTS #' = 'grts',
            'Visits' = 'visits',
            'Species' = 'species',
            'Mean disturbance' = 'mean_disturb'),
        colorNA = 'grey40',
        textNA = 'Not surveyed',
        clustering = FALSE) +
    
      tm_shape(park_point_sf_small) +
      tm_fill(
        col = 'species',
        style = 'cont',
        palette = 'magma',
        #breaks = c(0, 10, 20, 30, 40),
        id = 'point_name',
        popup.vars = 
          c('Point name' = 'point_name',
            'GRTS #' = 'grts',
            'Visits' = 'visits',
            'Species' = 'species',
            'Mean disturbance' = 'mean_disturb'),
        colorNA = 'grey40',
        textNA = 'Not surveyed',
        clustering = FALSE)

  tmap_leaflet(richness_map, mode = 'view')
  }
```

#### ANTI

```{r point_map_ANTI}
make_map('ANTI')
```

#### HAFE

```{r point_map_HAFE}
make_map('HAFE')
```

#### MANA

```{r point_map_MANA}
make_map('MANA')
```

#### MONO

```{r point_map_MONO}
make_map('MONO')
```

## All-time summary

This document summarizes the entire grassland bird dataset.

### Monitoring effort

### Survey conditions

### Bird detections

#### Top species by park

#### Species of concern over time

Occupancy of grassland species
Abundance of grassland species



