---
title: "Grassland field season summary"
date: "Generated `r Sys.Date()`"
output: html_document
params:
  year: 
    label: "Year"
    value: 2021
    input: slider
    min: 2014
    max: 2021
    step: 1
    sep: ""
---

```{r setup, include = FALSE}

# Note: this summarizer was used on "tidied" data.
# The script to tidy the data from the form I received it in can be found 
# in the script 1_import_bird_data.R

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
options(knitr.table.format = "html")
options(knitr.kable.NA = "")

# Load required packages

library(sf) # spatial data
library(tmap) # Interactive maps
library(tidyverse) # Data wrangling, plotting, and more
library(kableExtra) # Nice HTML tables
library(DT)

# Set plot theme

theme_set(theme_bw())

# Provide long names
parknames <-
  tibble(
    park = c("ANTI", "HAFE", "MANA", "MONO"),
    parkname = c(
      "Antietam National Battlefield", 
      "Harpers Ferry National Historical Park", 
      "Manassas National Battlefield Park", 
      "Monocacy National Battlefield"))
```

```{r eval = F}
# For testing
params <- list()
params$year <- 2021
```

**NPS Inventory and Monitoring, National Capital Region Network** 

Seasonal data summary tool created by Megan Massa, University of Maryland Center for Environmental Science, Appalachian Laboratory

<hr>

This tab summarizes survey efforts and results for grassland bird monitoring during the `r params$year` field season.

```{r data_import, include = FALSE}

# Import tidied data (see 1_import_bird_data.R)

read_rds("data/processed/birds.rds") |> 
  list2env(.GlobalEnv)

visits_yr <- filter(visits, year == params$year)

counts_yr <- filter(counts, visit_id %in% visits_yr$visit_id)

# Create summary: number of visits
n_visits <- 
  visits_yr |> 
  group_by(grts) |> 
  summarize(visits = length(unique(visit_id))) |> 
  group_by(visits) |> 
  summarize(n_points = n())

# Create variables for summary paragraph
total_pts <- 
  colSums(n_visits)[2]

three_visits <- 
  if(any(n_visits$visits == 3)) {filter(n_visits, visits == 3)[[2]]} else {0}

two_visits <- 
  if(any(n_visits$visits == 2)) {filter(n_visits, visits == 2)[[2]]} else {0}

one_visit <- 
  if(any(n_visits$visits == 1)) {filter(n_visits, visits == 1)[[2]]} else {0}

date_min <-
  min(visits$date) |> format("%d %b")

date_max <-
  max(visits$date) |> format("%d %b")

time_min <-
  round(abs(min(visits$start_sun)))

time_max <-
  round(abs(max(visits$start_sun)))
```

### Monitoring effort

The bird monitoring team conducted 
`r nrow(visits_yr)` 
surveys at 
`r total_pts` 
total points, of which 
`r three_visits`
`r if_else(three_visits == 1, "point", "points")`
received three visits,
`r two_visits` 
`r if_else(two_visits == 1, "point", "points")` 
received two visits, and
`r one_visit` 
`r if_else(one_visit == 1, "point", "points")` 
received only one visit. 
Surveys ran from `r date_min` to `r date_max` and were conducted from `r time_min` minutes before local sunrise to `r time_max` minutes (`r round(time_max / 60, 2)` 
hours) after local sunrise. 

```{r observer_table, fig.cap = paste("Number of surveys performed by each observer in", params$year)}
visits_yr |> 
  group_by(observer) |> 
  summarize(surveys = n()) |> 
  kable(col.names = c("Observer", "Surveys")) |> 
  kable_styling(full_width = FALSE)
```

### Survey conditions

The following graphs summarize environmental conditions during surveys. Dotted red lines indicate threshold recommended values according to survey protocol. Surveys past these lines were not conducted under advised conditions.

```{r survey_conditions1, fig.show = "hold", fig.asp = 0.25}

# Disturbance
ggplot() + 
geom_bar(
  data = visits_yr, 
  aes(x = disturbance)) +
scale_x_continuous(
  breaks = seq(
    from = 0, 
    to = max(visits$disturbance, na.rm = T), 
    by = 1))

# Wind
ggplot() + 
geom_bar(
  data = visits_yr, 
  aes(x = wind)) +
scale_x_continuous(
  breaks = seq(
    from = 0, 
    to = max(visits$wind, na.rm = T), 
    by = 1)) +
geom_vline(
  xintercept = 3.5, 
  color = "red", 
  linetype = "dashed")

# Sky
ggplot() + 
geom_bar(
  data = visits_yr, 
  aes(x = sky))

# Start time
ggplot(
  visits_yr, 
  aes(x = start_sun)) + 
geom_histogram(
  #stat = "density",
  binwidth = 10, 
  closed = "left") + 
geom_vline(
  xintercept = -20 - 10, 
  color = "red", 
  lty = 2) + 
geom_vline(
  xintercept = 300 - 10, 
  color = "red", 
  lty = 2) +
labs(x = "start time: minutes since sunrise")

# Date
ggplot(
  visits_yr, 
  aes(x = date)) + 
geom_histogram(binwidth = 1) 

# # Temperature
# ggplot(
#   visits_yr, 
#   aes(x = temperature)) + 
# geom_histogram(bins = 30) +
# labs(x = "temperature (C)")
# 
# # Humidity
# ggplot(
#   visits_yr, 
#   aes(x = humidity)) + 
# geom_histogram(bins = 30) +
# labs(x = "humidity (%)")
```

```{r survey_conditions2, fig.asp = 0.7}
# Temp vs humid
ggplot(
  visits_yr, 
  aes(x = humidity, y = temperature)) + 
geom_point() 
```


### Bird detections

```{r species_summary}
taxonomy <-
  read_rds("data/processed/taxonomy.rds") |> 
  mutate(
    tax_level = if_else(species == "ROPI", "species", tax_level),
    common_name = str_remove_all(common_name, " \\(Feral Pigeon\\)"),
    sci_name = str_remove_all(sci_name, " \\(Feral Pigeon\\)"))

species_list <-
  distinct(counts_yr, species) |> 
  left_join(taxonomy) |> 
  left_join(
    readxl::read_xlsx("data/raw/ACAD Global 2021.02.05.xlsx", 1) |> 
      select(
        common_name = `Common Name`, 
        habitat1 = `Primary Breeding Habitat`, 
        habitat2 = `Secondary Breeding Habitat`,
        iucn = `IUCN Red List 2018`,
        pif = `CI`),
    by = "common_name") |> 
  arrange(tax_order)

full_data <-
  counts_yr |> 
  left_join(visits) |> 
  left_join(points) |> 
  left_join(species_list)

grass_spp <- 
  species_list |> 
  filter(str_detect(habitat1, "Grassland") | str_detect(habitat2, "Grassland")) |> 
  pull(species)

# All obligate grassland species of temperate north america from Vickery et al. 1999
vickery <-
  c("NOHA","SWHA","FEHA","RLHA","STGR","GRPC","LEPC","MOPL","UPSA","LBCU","MAGO","BUOW","LEOW","SEOW","HOLA","SEWR","AMPI","SPPI","BACS","VESP","LABU","SAVS","GRSP","BAIS", "HESP","LCSP","TBLO","CCLO","DICL","BOBO","EAME","WEME")

# All facultative grassland species of temperate north america from Vickery et al. 1999
vickery2 <-
  c("AMBI","CAEG","TUVU","CANG","GADW","AMWI","MALL","BWTE","NSHO","NOPI","GWTE","AMKE","MERL","PEFA","PRFA","RNEP","WIPT","NOBO","YERA","SACR","WHCR","KILL","LEYE","WILL","LESA","WISN","WIPH","FRGU","MODO","BANO","CONI","COPO","SAPH","ATFL","CAKI","WEKI","EAKI","STFL","LOSH","NSHR","CHRA","EABL","WEBL","MOBL","COYE","CANT","RWSP","RCSP","CCSP","LASP","RWBL","BRBL","BRCO","BHCO")

vickery_spp <- filter(species_list, species %in% c(vickery, vickery2, grass_spp))$species

cbsd_spp <- filter(species_list, pif == "CBSD")$species

yellow_spp <- filter(species_list, pif == "Watch List - Yel-D")$species

park_summary <-
  full_data |> 
  mutate(
    grassland = if_else(species %in% c(vickery_spp, grass_spp), "yes", "no")) |> 
  group_by(park) |> 
  summarize(
    surveys = n_distinct(visit_id),
    points = n_distinct(grts),
    grass = n_distinct(species[grassland == "yes"]),
    species = n_distinct(species[tax_level == "species"]))

species_by_park <-
  full_data |> 
  select(park, grts, any_of(names(species_list))) |> 
  summarize(
    detections = n(), 
    points = n_distinct(grts),
    .by = c(species, park)) |> 
  left_join(species_list) |> 
  arrange(tax_order) |> 
  ungroup() |> 
  left_join(park_summary |> select(park, total_points = points)) |> 
  mutate(prop_occu = round(points/total_points, 2))
  
```

In `r params$year` there were `r filter(species_list, tax_level == "species") |> nrow()` bird species detected during grassland surveys. Observers detected `r n_distinct(c(grass_spp, vickery_spp))` grassland species within the battlefield parks, defined as species that are either obligate or facultative breeders in grasslands, as specified by Vickery (1999) and the Partners in Flight (PIF) Avian Conservation Assessment Database.  At grassland points observers detected `r length(cbsd_spp)` species ranked by PIF as Common Birds in Steep Decline and `r length(yellow_spp)` species placed on the PIF Yellow Watch List D for steep declines and major threats.

```{r park_table}
kable(
  park_summary,
  col.names = c("Park", "Surveys", "Points", "Grassland spp.", "Total spp.")) |> 
  kable_styling(full_width = FALSE)
```

#### Grassland species by park {.tabset}

The following tables show all species detected in the park this year, and notes those designated as obligate or facultative grassland species, along with all species of concern designated by Partners in Flight and whether they are facultative or obligate grassland species (Vickery 1999). Occupancy refers to the proportion of points within the park where the species was detected this year.

```{r specieslist}
concern_species <-
  species_by_park |> 
  mutate(grassland = case_when(
    species %in% vickery ~ "obligate",
    species %in% vickery2 ~ "facultative",
    TRUE ~ NA)) |> 
  select(
    park, species, common_name, sci_name, 
    tax_level, tax_order, detections, prop_occu, pif, grassland) |> 
  arrange(tax_order)

tables <-
  map(
    parknames$park,
    ~filter(
      concern_species,
      park == .x) |> 
      select(-park, -tax_level) |> 
      DT::datatable(
        colnames = c(
          "Alpha code",
          "Common name",
          "Scientific name",
          "Taxonomic order",
          "Detections",
          "Occupancy",
          "Watchlist",
          "Grassland status"))
    )

```

##### `r parknames$park[1]`

Grassland species detected at `r parknames$park[1]` in `r params$year`:

```{r}
tables[[1]]
```

##### `r parknames$park[2]`

Grassland species detected at `r parknames$park[2]` in `r params$year`:

```{r}
tables[[2]]
```

##### `r parknames$park[3]`

Grassland species detected at `r parknames$park[3]` in `r params$year`:

```{r}
tables[[1]]
```

##### `r parknames$park[4]`

Grassland species detected at `r parknames$park[4]` in `r params$year`:

```{r}
tables[[1]]
```

### Grassland species richness by point {.tabset}

```{r point_summary}
point_sf <-
  full_data |> 
    group_by(grts) |> 
    filter(species %in% c(grass_spp, vickery_spp)) |> 
    summarize(
      n_grass = length(unique(species)),
      grass_sp = paste0(unique(species), collapse = ", "),
      visits = length(unique(visit_id)),
      mean_disturb = mean(disturbance, na.rm = TRUE)) |> 
    left_join(points) |> 
    st_as_sf(coords = c('long', 'lat'), crs = 4326)
```

The total number of grassland species detected at each point in `r params$year` is mapped below.

```{r point_map_setup, include = FALSE}
park_lines <-
  st_read('data/processed/focal_parks.shp') |> 
    st_transform(crs = 4326) |> 
    st_make_valid()

make_map <- function(park){
  
  park_boundary <-
    filter(park_lines, UNIT_CODE == park) |> 
    st_transform(3857)
  
  park_point_sf <-
    filter(point_sf, park == park) |> 
    st_transform(3857) |> 
    st_buffer(dist = 100, nQuadSegs = 100)
  
  tmap_options(check.and.fix = TRUE)
  
  richness_map <- 
    tm_basemap(c("OpenTopoMap", "Esri.WorldImagery")) +
        
    # SAMPLING UNITS
    tm_shape(park_boundary) +
    tm_borders(
      col = 'black',
      lwd = 2.75) +
    
    tm_shape(park_point_sf) +
    tm_fill(
      col = 'n_grass',
      style = 'cont',
      palette = 'RdPu',
      id = 'point_name',
      alpha = 0.85,
      popup.vars = 
        c('Point name' = 'point_name',
          'GRTS #' = 'grts',
          'Visits' = 'visits',
          '# Grassland spp.' = 'n_grass',
          "Grassland spp." = "grass_sp"),
      colorNA = 'grey40',
      textNA = 'Not surveyed',
      clustering = FALSE) +
    
    tm_shape(park_point_sf) +
    tm_borders()

  tmap_leaflet(richness_map, mode = 'view')
  }
```

#### `r parknames$park[1]`

Map of grassland species detected at `r parknames$park[1]` in `r params$year`:

```{r}
make_map(parknames$park[1])
```

#### `r parknames$park[2]`

Map of grassland species detected at `r parknames$park[2]` in `r params$year`:

```{r}
make_map(parknames$park[2])
```

#### `r parknames$park[3]`

Map of grassland species detected at `r parknames$park[3]` in `r params$year`:

```{r}
make_map(parknames$park[3])
```

#### `r parknames$park[4]`

Map of grassland species detected at `r parknames$park[4]` in `r params$year`:

```{r}
make_map(parknames$park[4])
```

## All-time summary

This document summarizes the entire grassland bird dataset.

### Monitoring effort

```{r}
birds <- 
  counts |> 
  left_join(visits) |> 
  left_join(points) |> 
  left_join(species_list)

annual_effort <-
  birds |> 
  group_by(year, park) |> 
  summarize(
    n_surveys = length(unique(visit_id)),
    n_points = length(unique(grts)),
    n_spp = length(unique(species)))

annual_occ <-
  birds |> 
  summarize(
    n_present = length(unique(grts)),
    .by = c(year, park, species)) |> 
  ungroup() |> 
  complete(
    species, nesting(park, year),
    fill = list(n_present = 0)) |> 
  left_join(annual_effort) |> 
  mutate(prop_occ = round(n_present/n_points, 2)) |> 
  left_join(taxonomy)

pal <- palette.colors(n = 4, palette = "Okabe-Ito") |> unname()

annual_occ |> 
  filter(species %in% c("EAME", "GRSP", "HOLA")) |> 
  ggplot(aes(x = year, y = prop_occ, color = park)) +
    geom_line() +
    geom_point(aes(shape = park)) +
  scale_color_manual(values = pal) +
  facet_wrap(~common_name) +
  labs(y = "Park-wide occupancy", color = "", shape = "") +
  coord_cartesian(ylim = c(0, 1), xlim = c(2014, 2021)) +
  theme(
    strip.background = element_blank(),
    legend.position = "top" 
  )

annual_effort |> 
  ggplot(aes(x = year, y = n_surveys, color = park)) +
  geom_line() +
  geom_point(aes(shape = park)) +
  scale_color_manual(values = pal) +
  theme(legend.position = "top") +
  labs(y = "Total surveys", color = "", shape = "")

annual_effort |> 
  ggplot(aes(x = year, y = n_points, color = park)) +
  geom_line() +
  geom_point(aes(shape = park)) +
  scale_color_manual(values = pal) +
  theme(legend.position = "top") +
  labs(y = "Points surveyed", color = "", shape = "")

annual_effort |> 
  ggplot(aes(x = year, y = n_spp, color = park)) +
  geom_line() +
  geom_point(aes(shape = park)) +
  scale_color_manual(values = pal) +
  theme(legend.position = "top") +
  labs(y = "Total species detected", color = "", shape = "")

```